---
title: "How income varies across college major categories"
author: "Maurício Collaça"
date: "Nov 1st 2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The hyphotesis question

Is there an association between college major category and income?

## Requirements

    install.packages("devtools")
    devtools::install_github("jhudsl/collegeIncome")
    devtools::install_github("jhudsl/matahari")

```{r message=FALSE, warning=FALSE}
library(collegeIncome)
data(college)
library(matahari)
library(ggplot2);library(GGally);library(dplyr)
```

Custom functions
```{r}
wordwrap <- function(x, width = 10)
    lapply(strwrap(x, width = width, simplify = FALSE), paste, collapse="\n")
labelwrap <- function(x)
    gsub("_", " ", sub("perc_", "% ", x))
customtable <- function(x) 
    knitr::kable(x, col.names = labelwrap(names(x)), digits = 3)
pval <- function(x) {
    f <- summary(x)$f
    p <- pf(f[1], f[2], f[3], lower.tail = FALSE)
    attributes(p)<-NULL
    p
}
```

## Data validations

A codebook for the dataset is given below:

* rank: Rank by median earnings
* major_code: Major code
* major: Major description
* major_category: Category of major
* total: Total number of people with major
* sample_size: Sample size of full-time, year-round individuals used for income/earnings estimates: p25th, median, p75th
* p25th: 25th percentile of earnings
* median: Median earnings of full-time, year-round workers
* p75th: 75th percentile of earnings
* perc_men: % men with major (out of total)
* perc_women: % women with major (out of total)
* perc_employed: % employed (out of total)
* perc_employed_fulltime: % employed 35 hours or more (out of employed)
* perc_employed_parttime: % employed less than 35 hours (out of employed)
* perc_employed_fulltime_yearround: % employed at least 50 weeks and at least 35 hours (out of employed and full-time)
* perc_unemployed: % unemployed (out of employed)
* perc_college_jobs: % with job requiring a college degree (out of employed)
* perc_non_college_jobs: % with job not requiring a college degree (out of employed)
* perc_low_wage_jobs: % in low-wage service jobs (out of total)

### Generalized pairs plot of all numeric variables

There are some expected relatioships:

* The sample size and total number of people with major are highly positive correlated.
* The percent of woman and men are perfectly negative correlated.
* The percent of employed and unemployed are perfectly negative correlated.

There are also unexpected relationships:

* The rank is not strongly related with anything
* The percent of employed fulltime can't be correlated, suggesting that Infinite values happened.
* The percent of college jobs and non college jobs are not perfectly negative correlated which denotes miscalculations.

```{r fig.height=20, fig.width=20, message=FALSE, warning=FALSE, cache=TRUE}
college %>%
    select(everything(), -c(2:4)) %>%
    ggpairs(cardinality_threshold = 16, columnLabels = labelwrap(names(.)))
```

### Each observation is a major
```{r}
nrow(college)==length(unique(college$major))
```

### Rank uniformity and discreteness
```{r}
all(sort(college$rank) == 1:nrow(college))
```

### Incomplete cases
```{r}
NACols <- names(which(colSums(is.na(college)) > 0))
college %>%
    filter(!complete.cases(.)) %>%
    select(major_category, major_code, one_of(NACols)) %>%
    customtable
```

### Observations with invalid percents

A helper function to check whether the value is a wrong percent, assuming a range from 0 to 1, inclusive.
```{r}
is.wrong <- function(x, min=0, max=1) is.na(x) | is.infinite(x) | (is.numeric(x) & (x < min | x > max))
```

Which columns have wrong values
```{r}
(wrongCols <- names(which(colSums(sapply(college[, grep("^perc_",names(college), value = TRUE)], is.wrong))>0)))
```

Which rows have wrong values
```{r}
wrongRows <- apply(sapply(wrongCols, function(col) is.wrong(college[, col])), 1, any)
college %>%
    filter(wrongRows) %>%
    select(rank, major_category, major_code, one_of(wrongCols)) %>%
    customtable
```

Invalid values such as `NaN` (not a number), `Inf` (infinite) and `NA` (missing) foster problems and bias in calculations.  It's a best practice to get rid of them, as much as it makes sense, documenting those transformations for reproducibility purposes.

Also, percents must not be outside the range 0 to 1.

There's no raw data to recalculate the underlying percents.

As the hypothesis question is directly to the rank and not the percents, and, the ranks are perfectly uniform, it's assumed that a good strategy is not to remove incomplete observations but fix the wrong percents by their major category mean, or prefferably median, avoiding bias as and if the percents were taken as regression terms.

## Data transformations

### Category abbreviation

Abbreviate major category to fit in tiny spaces, e.g. axis labels report columns
```{r}
college <- college %>%
    mutate(category = factor(abbreviate(major_category, minlength=3)))

college %>%
    select(category, major_category) %>%
    distinct %>%
    customtable
```

### Imputing median values

As the percents are originated from median statistics and in order to minimize bias, it's assumed the median is a better imputing value than the mean.
```{r}
college <- college
for(col in seq_along(wrongCols)) {
    wrongColRows <- is.wrong(college[, wrongCols[col]])
    college[wrongColRows, wrongCols[col]] <- 
        median(college[!wrongColRows, wrongCols[col]])
}
```

Rows after imputation
```{r}
college %>%
    filter(wrongRows) %>%
    select(rank, major_category, major_code, one_of(wrongCols)) %>%
    customtable
```

## Exploratory Data Analysis

As stated in the codebook, the rank is calculated from median earnings of full-time, year-round workers, understood as the rank is the response variable and its main predictor is the median.

he following percents pairs are understood as complementary, i.e., each sum should result in 1 (100%):

    perc_men: % men with major (out of total)
    perc_women: % women with major (out of total)
    
    perc_employed: % employed (out of total)
    perc_unemployed: % unemployed (out of employed)
    
    perc_employed_fulltime: % employed 35 hours or more (out of employed)
    perc_employed_parttime: % employed less than 35 hours (out of employed)

    perc_college_jobs: % with job requiring a college degree (out of employed)
    perc_non_college_jobs: % with job not requiring a college degree (out of employed)

However, the following percents are understood as standalone:

    perc_employed_fulltime_yearround: % employed at least 50 weeks and at least 35 hours (out of employed and full-time)
    perc_low_wage_jobs: % in low-wage service jobs (out of total)

These assumptions drive the following explorations.

### Generalized pairs plot of all numeric variables after imputing

After imputing data, the previoulsy expected relatioships remain:

* The sample size and total number of people with major are highly positive correlated .
* The percent of woman and men are perfectly negative correlated.
* The percent of employed and unemployed are perfectly negative correlated.

After imputing data, the previously unexpected relationships changed but still remain:

* The rank is not strongly related with any variable.
* The percent of employed fulltime now can be correlated, however, it's not perfectly negative correlated with the percent of unemployed.
* The percent of college jobs and non college jobs are not perfectly negative correlated as they weren't before.

```{r fig.height=20, fig.width=20, message=FALSE, warning=FALSE, cache=TRUE}
college %>%
    select(everything(), -c(2:4,20)) %>%
    ggpairs(cardinality_threshold = 16, columnLabels = labelwrap(names(.)))
```

### Ranking reproducibility

Although the codebook states that the rank is by median earnings, it was not possible to reproduce it with the function `rank()`.  It's found even an unexpected negative correlation between them. The analysis will continue without knowning the ranking process.

```{r}
ggplot(college, aes(x=rank, y=rank(median))) +
    geom_point(aes(color=category)) +
    geom_smooth(method="lm")
```

### Rank by major category

The box plot shows the best ranked majors are from Engineering and the worst from Psychology & Social Work.

```{r fig.height=3, fig.width=9}
ggplot(college, aes(y=rank, x=category, fill=category)) +
    geom_boxplot(show.legend = FALSE)
```

The violin plot overlaid with a dot plot give a better view of the rank distribution by category. It suggests:

* Engineering (Eng) ranks are more concentrated above 25, followed by Computers & Mathematics (C&M) more concentrated around 50..
* Communications & Journalism (C&J) ranks are concentrated slightly above 100.
* Categories "Art" and "Psychology & Social Work" have ranks more concentrated around 150.
* The Interdisciplinary category has only one major, ranked bellow 100.
* The other categories not mentioned above are more uniformly distributed but worth metion that the categories Business (Bsn) and Physical Sciences (PhS) have each one a major that is around the overall top ones.

```{r fig.height=3, fig.width=9}
ggplot(college, aes(y=rank, x=category, fill=category)) +
    geom_violin(show.legend = FALSE) +
    geom_dotplot(binaxis='y', stackdir='center', fill="black",
                 dotsize=.6, binwidth = 5)
```

### Median earning distribution

As stated in the codebook, the rank is calculated from median earnings of full-time, year-round workers, therefore, this variable should be investigated.

Comparing the distribution kernel density and the normal curve one can see that the distribution of the median earning of each major is not normally distributed.

```{r}
x <- college$median
ggplot(college, aes(x=median), size=1) +
    geom_density(aes(color="Kernel"), fill="dark red", alpha=0.2) +
    stat_function(aes(color="Normal"), fun=dnorm, args=list(mean=mean(x), sd=sd(x))) +
    scale_color_manual("Density", values = c(Kernel="dark red", Normal="blue")) +
    geom_vline(aes(xintercept = median(x), linetype="Median")) +
    geom_vline(aes(xintercept = mean(x), linetype="Mean")) +
    scale_linetype_manual("Centrality", values = c(Median=1, Mean=2))
```

### Median earning by major category

As stated in the codebook, the Rank is calculated from median earnings of full-time, year-round workers, therefore, the distribution of median income by category is an interesting exploration.

There are very potential influence outliers that deservers investigation.

```{r fig.height=4, fig.width=9}
ggplot(college, aes(y=median, x=category, fill=category)) +
    geom_boxplot(show.legend = FALSE)
```

```{r fig.height=4, fig.width=9}
ggplot(college, aes(y=median, x=category, fill=category)) +
    geom_violin() +
    geom_dotplot(binaxis='y', stackdir='center', fill="black",
                 dotsize=1.5, binwidth = 1000) +
    theme(legend.position = "none")
```

### Rank by median incoming

It's already detected the presence of outliers.

As stated in the codebook, the Rank is predicted by median earnings of full-time, year-round workers, therefore, it's important to understand the relationship between rank and median earning.

The overall regression line (black) suggests the higher the median earning the higher (low order) the rank.

The slope sign inversions suggests further investigation.

```{r}
college %>% 
    ggplot(aes(y=rank, x=median, color=category)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(aes(color=NULL), color="black", method = "lm")
```

One can see the top 10 median earnings and respective ranks. The major "Miscellaneous Business & Medical Administration" have much higher median earning than the other majors but figures at 63rd in the rank while the major "Chemical Engineering" figures at 5th in the rank but with a much lower median earning.

```{r}
college %>%
    arrange(desc(median)) %>%
    slice(1:10) %>%
    select(category, major_code, major, total, sample_size, median, rank) %>%
    customtable()
```

Without the raw data, one can't ensure wheter outliers are results from spurious or real process, however, residual analysis is recommended in order to measure influence.

### Rank by median earning conditioned by correlation sign

To eventually avoid or at least isolate a Sympson paradox, the relationship between the rank and the median earning is conditioned by the correlation sign.  It's important to remind that a negative correlation, in this case, means a higher rank (low order) position.

The right-side overall regression line (black) seems to better isolate the Sympson Paradox between 5 categories and still shows the previously mentioned influential values.  Within those categories, the more the median incomig increases, the lower is the rank (high order), i.e., those categories have unexpected positive correlations between the rank and the median however altogether result in the expected negative correlation.

```{r}
college %>%
    group_by(category) %>%
    mutate(correlation_signal = sign(cor(rank, median))) %>%
    ungroup %>%
    filter(!is.na(correlation_signal)) %>%
    ggplot(aes(y=rank, x=median, color=category)) +
    facet_grid(. ~ correlation_signal) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(aes(color=NULL), color="black", method = "lm")
```

Note: the category "Int" (Interdisciplinary) had only one observation thus can't be correlated, thus removed from the plot above with `filter(!is.na(correlation_signal))`.

### Outliers

Leverage measures (hat values) can be useful for diagnosing data entry errors and points that have a high potential for influence.

Influence measures explains how removing points impact a particular aspect of the model.

#### Residual plots 

In the first plot there's some kind of undesired systematic pattern and three outlying observations.

In the second plot there is a violation in the normality assumption of the residuals.

```{r fig.width=9, fig.height=10}
par(mfrow=c(3,2))
plot(lm(rank ~ median, college),
     which = 1:6, cex.id = 1, col="dark gray")
```

#### Influence measures

```{r}
rankByMedianModel <- lm(rank ~ median, college)
rankByMedianMeasures <- influence.measures(rankByMedianModel)
summary(rankByMedianMeasures)
```

#### Influential observations
```{r}
influential <- unname(apply(rankByMedianMeasures$is.inf, 1, any))
college %>%
    select(category, major_code, major, total, sample_size, median, rank) %>%
    filter(influential) %>% customtable()
```

#### What-if removing influential observations

```{r}
which(influential)
```

Removing the influential observations avoided the Sysmpson paradox in the correlation sign which suggests an improved model fit.

```{r}
college %>%
    filter(!influential) %>%
    group_by(category) %>%
    mutate(correlation_signal = sign(cor(rank, median))) %>%
    ungroup %>%
    filter(!is.na(correlation_signal)) %>%
    ggplot(aes(y=rank, x=median, color=category)) +
    facet_grid(. ~ correlation_signal) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(aes(color=NULL), color="black", method = "lm")
```

## Regression inference

### The initial hypothesis model
```{r comment=""}
fit <- lm(rank ~ category, college)
anova(fit)
```

### Removing intercept yields higher F-statistic 
```{r comment=""}
fit <- lm(rank ~ category - 1, college)
anova(fit)
```

### Residual plot of the best model so far
```{r fig.width=9, fig.height=10}
par(mfrow=c(3,2))
plot(fit, which = 1:6, cex.id = 1, col="dark gray")
```

This colored point version of the Residuals versus Fit plot helps identify patterns in categories.

```{r}
ggplot(college, mapping = aes(x=predict(fit), y=resid(fit), color=category)) +
    geom_point() +
    geom_smooth(aes(color=NULL), method="lm")
```

### Influence measurements
```{r}
measures <- influence.measures(fit)
summary(measures)
```

### Influential observations
```{r}
influential <- unname(apply(measures$is.inf, 1, any))
college %>%
    select(category, major_code, major, total, sample_size, median, rank) %>%
    filter(influential) %>% customtable()
```

### Nested models

```{r comment=""}
fit1 <- lm(rank ~ 0 + category, college)
fit2 <- lm(rank ~ 0 + category + I(college$perc_men > .5), college)
fit3 <- lm(rank ~ 0 + category + I(college$perc_women > .5), college)
fit4 <- lm(rank ~ 0 + category + I(perc_employed > .5), college)
fit5 <- lm(rank ~ 0 + category + I(perc_unemployed > .5), college)
fit6 <- lm(rank ~ 0 + category + I(perc_employed_fulltime > .5), college)
fit7 <- lm(rank ~ 0 + category + I(perc_employed_parttime > .5), college)
fit8 <- lm(rank ~ 0 + category + I(perc_employed_fulltime_yearround > .5), college)
fit9 <- lm(rank ~ 0 + category + I(!perc_employed_fulltime_yearround > .5), college)
fit10 <- lm(rank ~ 0 + category + I(perc_college_jobs > .5), college)
fit11 <- lm(rank ~ 0 + category + I(perc_non_college_jobs > .5), college)
fit12 <- lm(rank ~ 0 + category + I(perc_low_wage_jobs > .5), college)
fit13 <- lm(rank ~ 0 + category + I(!perc_low_wage_jobs > .5), college)
fit14 <- lm(rank ~ 0 + category * I(perc_non_college_jobs > .5), college)
fit15 <- lm(rank ~ 0 + category : I(perc_non_college_jobs > .5), college)
```

From all these models, the first is the best.  The model 10 was the one with the least worst result.

```{r}
anova(fit1)
anova(fit1, fit10)
```

## Appendix

### All the nested models ANOVA results:
```{r comment=""}
anova(fit1, fit2)
anova(fit1, fit3)
anova(fit1, fit4)
anova(fit1, fit5)
anova(fit1, fit6)
anova(fit1, fit7)
anova(fit1, fit8)
anova(fit1, fit9)
anova(fit1, fit10) #least worst
anova(fit1, fit11)
anova(fit1, fit12)
anova(fit1, fit13)
anova(fit1, fit14)
anova(fit1, fit15)
```
